---
title: "Preprocessing Meta Data"
author: "Zhongyi (James) Guo"
date: "`r Sys.Date()`"
output: html_document
---

We extracted metadata (covariates) using `Table Exporter`. This file contains metadata cleaning and exploratory data analysis.

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(readr)
library(scales)
setwd("~/ukb-nmr/scripts/data_prep")
```

## Data Cleaning

```{r import_metadata, message=FALSE}
metadata <- read_csv('~/pca_metabolomics/table_exporter/output/metadata__participant.csv')
dim(metadata)
```

There were 228,973 participants and 13 columns (including `eid`).

```{r map_field_titles, message=FALSE}
# remove instance number and "p"
fields_of_interest <- sub("^p", "", sub("_.*", "", colnames(metadata)[-1]))

# extract field titles from field dictionary
field_dictionary <- read_tsv('~/pca_metabolomics/Showcase metadata/field.tsv')
field_titles <- field_dictionary |> 
  filter(field_id %in% fields_of_interest) |> 
  slice(match(fields_of_interest, field_id)) |>
  pull(title)

colnames(metadata) <- c('eid', field_titles)
```

```{r preview_final_metadata}
metadata
```

## Exploratory Data Analysis

NA rate:

```{r NA_rate_of_all_columns}
 colMeans(is.na(metadata))
```

```{r age_at_recruitment_distribution}
boxplot(metadata$`Age at recruitment`, main = 'Histogram of Age at Recruitment', xlab = 'Age at Recruitment')
```

```{r ethnic_background_pie_chart}
metadata |> 
  group_by(`Ethnic background`) |> 
  summarise(count = n()) |>
  ggplot(aes(x = "", y = count, fill = `Ethnic background`)) +
  geom_bar(stat = "identity", width = 1) +
    coord_polar(theta = "y") +
    labs(title = "Pie Chart of Ethnic Background") +
    theme_minimal() +
    theme(axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(),
          legend.text = element_text(size = 7),
          legend.key.size = unit(0.5, "cm"))
```

```{r father_with_prostate_cancer_percentage}
father_with_prostate_cancer <- metadata[grepl("prostate cancer", metadata$`Illnesses of father`, ignore.case = TRUE), ]
nrow(father_with_prostate_cancer)
```

15,114 participants reported that their fathers had prostate cancer, which accounted for `r 15114 / nrow(metadata) * 100`% of the total participants.

```{r siblings_with_prostate_cancer_percentage}
siblings_with_prostate_cancer <- metadata[grepl("prostate cancer", metadata$`Illnesses of siblings`, ignore.case = TRUE), ]
nrow(siblings_with_prostate_cancer)
```

2,928 participants reported that their siblings had prostate cancer, which accounted for `r 15114 / nrow(metadata) * 100`% of the total participants.

```{r bmi_distribution}
boxplot(metadata$`Body mass index (BMI)`, main = 'Histogram of BMI', xlab = 'BMI')
```

```{r time_blood_sample_collected_scatterplot}
as.Date(metadata$`Time blood sample collected`) |>
  as.data.frame() |>
  na.omit() |>
  setNames("date") |>
  group_by(date) |>
  summarise(count = n()) |>
  ggplot(aes(x = date, y = count)) +
  geom_point() +
  ggtitle('Number of Blood Samples Collected Per Day') +
  scale_x_date(
    date_breaks = "1 month",  
    date_labels = "%Y-%m-%d"
  ) +
  ylim(0, 400) + 
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)  # Rotate x-axis labels
  )
```

```{r fasting_time_distribution}
boxplot(metadata$`Fasting time`, main = 'Histogram of Fasting time', xlab = 'Fasting time')
```

```{r townsend_deprivation_index_at_recruitment_distribution}
boxplot(metadata$`Townsend deprivation index at recruitment`, main = 'Histogram of Townsend deprivation index at recruitment', xlab = 'Townsend deprivation index at recruitment')
```

```{r ever_smoked_pie_chart}
metadata |> 
  group_by(`Ever smoked`) |> 
  summarise(count = n()) |>
  ggplot(aes(x = "", y = count, fill = `Ever smoked`)) +
  geom_bar(stat = "identity", width = 1) +
    coord_polar(theta = "y") +
    labs(title = "Pie Chart of Ever Smoked") +
    theme_minimal() +
    theme(axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank())
```

```{r cholesterol_distribution}
boxplot(metadata$`Cholesterol`, main = 'Histogram of Cholesterol', xlab = 'Cholesterol')
```

```{r hdl_cholesterol_distribution}
boxplot(metadata$`HDL cholesterol`, main = 'Histogram of HDL cholesterol', xlab = 'HDL cholesterol')
```

## Save Cleaned Meta Data

```{r save_metadata}
write.csv(metadata, '~/pca_metabolomics/cleaned_data/metadata_cleaned.csv', row.names = FALSE)
```

```{r sessionInfo}
sessionInfo()
```
