---
title: "Preprocessing NMR Processing"
author: "Zhongyi (James) Guo"
date: "`r Sys.Date()`"
output: html_document
---

We extracted NMR processing using `Table Exporter`. This file contains NMR processing data cleaning and exploratory data analysis.

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(readr)
setwd("~/ukb-nmr/scripts/data_prep")
```

## Data Cleaning

```{r import_nmr_processing, warning=FALSE, message=FALSE}
nmr_processing <- read_csv('~/pca_metabolomics/table_exporter/output/nmr_processing_participant.csv')
dim(nmr_processing)
```

There were 228,973 participants and 8 columns (including `eid`).

```{r map_field_titles, message=FALSE, warning=FALSE}
# remove instance 1 columns
nmr_processing <- nmr_processing |>
  select(-matches("i1$"))

# remove instance number and "p"
fields_of_interest <- sub("^p", "", sub("_.*", "", colnames(nmr_processing)[-1]))

# extract field titles from field dictionary
field_dictionary <- read_tsv('~/pca_metabolomics/Showcase metadata/field.tsv')
field_titles <- field_dictionary |> 
  filter(field_id %in% fields_of_interest) |> 
  slice(match(fields_of_interest, field_id)) |>
  pull(title)

colnames(nmr_processing) <- c('eid', field_titles)
```

```{r preview_nmr_processing}
nmr_processing
```

## Exploratory Data Analysis

NA rate:

```{r NA_rate_of_all_columns}
 colMeans(is.na(nmr_processing))
```

```{r prostate_cancer_percentage}
with_prostate_cancer <- nmr_processing[grepl("C61", nmr_processing$`Type of cancer: ICD10`, ignore.case = FALSE), ]
nrow(with_prostate_cancer)
```

13,242 participants had prostate cancer, which accounted for `r 13242 / nrow(nmr_processing) * 100`% of the total participants.

```{r control_percentage}
icd10_na <- nmr_processing[is.na(nmr_processing$`Type of cancer: ICD10`), ]
nrow(icd10_na)
```

173,598 had no cancer, which accounted for `r 173598 / nrow(nmr_processing) * 100`% of the total participants.

```{r high_pyruvate_percentage}
unique(nmr_processing$`High Pyruvate`)
```

No values for the `High Pyruvate` column.

```{r low_protein_percentage}
table(nmr_processing$`Low Protein`)
```

853 had low protein.

```{r low_glucose_percentage}
unique(nmr_processing$`Low Glucose`)
```

No values for the `Low Glucose` column.

```{r high_lactage_percentage}
table(nmr_processing$`High Lactate`)
```

21,663 had high lactate.

## Save Cleaned NMR Processing

```{r save_nmr_processing}
write.csv(nmr_processing, '~/pca_metabolomics/cleaned_data/nmr_processing_cleaned.csv', row.names = FALSE)
```

```{r sessionInfo}
sessionInfo()
```
